Geometric models reveal behavioral and neural signatures of transforming experiences into memories

### 2020-10-16 paper/main.tex
updated supp fig numbers
added caption for new supplemental figure, formatting edits
udpated notes on formatting checklist, minor edits to main text

\newcommand{\topicopt}{1}
\newcommand{\topics}{2}
\newcommand{\featureimportance}{3}
\newcommand{\corrmats}{4}
\newcommand{\matchmats}{5}
\newcommand{\kopt}{6}
\newcommand{\arrows}{7}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/list_learning}
  \caption{\small \textbf{Naturalistic extensions of classic list-learning memory analyses.} \textbf{A.} The probability of first recall as a function of the serial position of the event in the episode. \textbf{B}.  The probability of recalling each event, conditioned on having most recently recalled the event \textit{lag} events away in the episode.  \textbf{C.} The proportion of participants who recalled each event, as a function of the serial position of the events in the episode.  All panels: error ribbons denote bootstrap-estimated standard error of the mean.}
  \label{fig:list-learning}
\end{figure}

Clustering scores are often used by memory researchers to characterize how people organize their memories of words on a studied list~\citep{PolyEtal09}.  We defined analogous measures to characterize how participants organized their memories for episodic events (see \textit{Methods} for details).  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed episode events tended to appear nearby in participants' recall event sequences (mean clustering score: 0.732, SEM: 0.033).  Participants with higher temporal clustering scores tended to exhibit better overall memory for the episode, according to both Chen et al. (2017)~\cite{ChenEtal17}'s hand-counted numbers of recalled scenes from the episode (Pearson's $r(15) = 0.49,~p = 0.046,~95\%~CI~[0.25, 0.76]$) and the numbers of episode events that best-matched at least one recall event (i.e., model-estimated number of events recalled; Pearson's $r(15) = 0.59,~p = 0.013,~95\%~CI~[0.31, 0.80]$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity~\citep{MannKaha12}.  We found that participants tended to recall semantically similar episode events together (mean clustering score: 0.650, SEM: 0.032), and that semantic clustering score was also related to both hand-counted  (Pearson's $r(15) = 0.65,~p = 0.004,~95\%~CI~[0.31, 0.85]$) and model-estimated (Pearson's $r(15) = 0.58,~p = 0.015,~95\%~CI~[0.10, 0.83]$) numbers of recalled events.

@@

In addition to examining how precisely and distinctively participants recalled individual events, one may also use these metrics to summarize each participant's performance by averaging across a participant's event-wise precision or distinctiveness scores.  This enables us to quantify how precisely a participant tended to recall subtle within-event details, as well as how specific (distinctive) those details were to individual events from the episode.  Participants' average precision and distinctiveness scores were strongly correlated ($r(15) = 0.90,~p < 0.001,~95\%~CI~[0.66, 0.96]$).  This indicates that participants who tended to precisely recount low-level details of episode events also tended to do so in an event-specific way (e.g., as opposed to detailing recurring themes that were present in most or all episode events; this behavior would have resulted in high precision but low distinctiveness).  We found that, across participants, higher precision scores were positively correlated with the numbers of both hand-annotated scenes ($r(15) = 0.60,~p = 0.010,~95\%~CI~[0.02, 0.83]$) and model-estimated events ($r(15) = 0.90,~p < 0.001,~95\%~CI~[0.54, 0.96]$) that participants recalled.  Participants' average distinctiveness scores were also marginally correlated with both the hand-annotated ($r(15) = 0.45,~p = 0.068,~95\%~CI~[-0.21, 0.79]$) and model-estimated ($r(15) = 0.71,~p = 0.001,~95\%~CI~[-0.07, 0.90]$) numbers of recalled events.

@@

\subsection*{Statistics}
All statistical tests performed in the behavioral analyses were two-sided.  All statistical tests performed in the neural data analyses were two-sided, except for the permutation-based thresholding, which was one-sided.  In this case, we were specifically interested in identifying voxels whose activation time series reflected the temporal structure of the episode and recall topic proportions matrices to a greater extent than that of the phase-shifted matrices.

@@

Perhaps the simplest and most widely used measure of memory performance is ``accuracy''---i.e., the proportion of studied (experienced) items (in this case, episode events) that the participant later remembered.  Chen et al.~(2017)~\cite{ChenEtal17} used this method to rate each participant's memory quality by computing the proportion of (50 manually identified) scenes mentioned in their recall.  We found a strong across-participants correlation between these independent ratings and the proportion of 30 HMM-identified episode events matched to participants' recalls (Pearson's $r(15) = 0.71, p = 0.002,~95\%~CI~[0.39, 0.88]$).  We further considered a number of more nuanced memory performance measures that are typically associated with list-learning studies.  We also provide a software package, \texttt{Quail}, for carrying out these analyses~\citep{HeusEtal17b}.

>>>>

\newcommand{\topicopt}{1}
\newcommand{\topics}{2}
\newcommand{\featureimportance}{3}
\newcommand{\corrmats}{5}
\newcommand{\matchmats}{6}
\newcommand{\kopt}{7}
\newcommand{\arrows}{4}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/list_learning}
  \caption{\small \textbf{Naturalistic extensions of classic list-learning memory analyses.} \textbf{A.} The probability of first recall as a function of the serial position of the event in the episode. \textbf{B}.  The probability of recalling each event, conditioned on having most recently recalled the event \textit{lag} events away in the episode.  \textbf{C.} The proportion of participants who recalled each event, as a function of the serial position of the events in the episode.  All panels: error ribbons denote the bootstrap-estimated 95\% confidence interval.}
  \label{fig:list-learning}
\end{figure}

Clustering scores are often used by memory researchers to characterize how people organize their memories of words on a studied list~\citep{PolyEtal09}.  We defined analogous measures to characterize how participants organized their memories for episodic events (see \textit{Methods} for details).  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed episode events tended to appear nearby in participants' recall event sequences (mean clustering score: 0.732, SEM: 0.033).  Participants with higher temporal clustering scores tended to exhibit better overall memory for the episode, according to both Chen et al. (2017)~\cite{ChenEtal17}'s hand-counted numbers of recalled scenes from the episode (Pearson's $r(15) = 0.49,~p = 0.046,~95\%~\mathrm{CI} = [0.25, 0.76]$) and the numbers of episode events that best-matched at least one recall event (i.e., model-estimated number of events recalled; Pearson's $r(15) = 0.59,~p = 0.013,~95\%~\mathrm{CI} = [0.31, 0.80]$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity~\citep{MannKaha12}.  We found that participants tended to recall semantically similar episode events together (mean clustering score: 0.650, SEM: 0.032), and that semantic clustering score was also related to both hand-counted  (Pearson's $r(15) = 0.65,~p = 0.004,~95\%~\mathrm{CI} = [0.31, 0.85]$) and model-estimated (Pearson's $r(15) = 0.58,~p = 0.015,~95\%~\mathrm{CI} = [0.10, 0.83]$) numbers of recalled events.

@@

In addition to examining how precisely and distinctively participants recalled individual events, one may also use these metrics to summarize each participant's performance by averaging across a participant's event-wise precision or distinctiveness scores.  This enables us to quantify how precisely a participant tended to recall subtle within-event details, as well as how specific (distinctive) those details were to individual events from the episode.  Participants' average precision and distinctiveness scores were strongly correlated ($r(15) = 0.90,~p < 0.001,~95\%~\mathrm{CI} = [0.66, 0.96]$).  This indicates that participants who tended to precisely recount low-level details of episode events also tended to do so in an event-specific way (e.g., as opposed to detailing recurring themes that were present in most or all episode events; this behavior would have resulted in high precision but low distinctiveness).  We found that, across participants, higher precision scores were positively correlated with the numbers of both hand-annotated scenes ($r(15) = 0.60,~p = 0.010,~95\%~\mathrm{CI} = [0.02, 0.83]$) and model-estimated events ($r(15) = 0.90,~p < 0.001,~95\%~\mathrm{CI} = [0.54, 0.96]$) that participants recalled.  Participants' average distinctiveness scores were also marginally correlated with both the hand-annotated ($r(15) = 0.45,~p = 0.068,~95\%~\mathrm{CI} = [-0.21, 0.79]$) and model-estimated ($r(15) = 0.71,~p = 0.001,~95\%~\mathrm{CI} = [-0.07, 0.90]$) numbers of recalled events.

@@

\subsection*{Statistics}
All statistical tests performed in the behavioral analyses were two-sided.  All statistical tests performed in the neural data analyses were two-sided, except for the permutation-based thresholding, which was one-sided.  In this case, we were specifically interested in identifying voxels whose activation time series reflected the temporal structure of the episode and recall topic proportions matrices to a greater extent than that of the phase-shifted matrices.  The 95\% confidence intervals we reported for each correlation were estimated by generating 10000 ``bootstrap'' distributions of correlation coefficients by sampling (with replacement) from the observed data.

@@

Perhaps the simplest and most widely used measure of memory performance is ``accuracy''---i.e., the proportion of studied (experienced) items (in this case, episode events) that the participant later remembered.  Chen et al.~(2017)~\cite{ChenEtal17} used this method to rate each participant's memory quality by computing the proportion of (50 manually identified) scenes mentioned in their recall.  We found a strong across-participants correlation between these independent ratings and the proportion of 30 HMM-identified episode events matched to participants' recalls (Pearson's $r(15) = 0.71, p = 0.002,~95\%~\mathrm{CI} = [0.39, 0.88]$).  We further considered a number of more nuanced memory performance measures that are typically associated with list-learning studies.  We also provide a software package, \texttt{Quail}, for carrying out these analyses~\citep{HeusEtal17b}.
